Design and Implementation of\\Anglican Probabilistic Programming Language

### 2016-09-01 paper/paper.tex
typo fixed
added worksheet
in good progres

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
% Although these generic techniques are not always as statistically
% efficient as techniques that take advantage of model-specific
% optimizations, probabilistic programming makes it easier to optimize
% models for a specific application in a manner that is efficient in
% terms of the dimensionality of its latent variables.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model, however
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved 
memoized functions. In Anglican, one might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

>>>>

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model, however
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

### 2016-09-01 slides/slides.tex
in good progres

\begin{frame}{Anglican and Clojure}
    On language on top of (or besides) another:
    \begin{itemize}
        \item Interpeter.
        \item Source-to-source compiler.
        \item Integrated language. $\checkmark$
    \end{itemize}
    \pause
    \bigskip
    Integrated language
    \begin{itemize}
        \item either extends the syntax of the host language;
        \item or changes the semantics of the host language
            constructs $\checkmark$.
    \end{itemize}
    \pause
    \bigskip
    Anglican is 
    \begin{itemize}
        \item Integrated with Clojure.
        \item Shares syntax.
        \item Alters operational semantics.
    \end{itemize}
\end{frame}

@@

\begin{frame}{Managing stack size}
\end{frame}

\begin{frame}{Probabilistic forms}
\end{frame}

\begin{frame}{Memoization}
\end{frame}

>>>>

\begin{frame}{Anglican and Clojure}
    One language on top of (or besides) another:
    \begin{itemize}
        \item Interpeter.
        \item Source-to-source compiler.
        \item Integrated language. $\checkmark$
    \end{itemize}
    \pause
    \bigskip
    Integrated language
    \begin{itemize}
        \item either extends the syntax of the host language;
        \item or changes the semantics of the host language
            constructs $\checkmark$.
    \end{itemize}
    \pause
    \bigskip
    Anglican is 
    \begin{itemize}
        \item Integrated with Clojure.
        \item Shares syntax.
        \item Alters operational semantics.
    \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{Probabilistic forms}
    Two \textbf{probabilistic} forms:
    \begin{itemize}
        \item \texttt{sample} --- draws a value from a
            distribution;
        \item \texttt{observe} --- conditions the \textit{a
            posteriori} distribution by observing a value from a
            distribution.
    \end{itemize}
    \bigskip
    This is where the inference algorithm steps in:
\begin{minted}[fontsize=\small]{clojure}
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)

=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Memoization}
Memoization of random choice:
\begin{itemize}
    \item A person has a random eye color.
    \item But the \textit{same} person has a fixed eye color.
\end{itemize}
\begin{minted}[linenos, fontsize=\small]{clojure}
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{minted}
\end{frame}

\begin{frame}[fragile]{Compiling a memoized function}
    \vspace{-2em}
    \begin{minipage}{0.38\textwidth}
\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
(mem (fn [person] ...))
\end{minted}
    \end{minipage}
    \begin{minipage}{0.52\textwidth}
    \vspace{2em}
\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
(fn []
 (cont
  ;; every memoization gets a unique key  
  (let [M23145 (gensym "M")]
   (fn [C23144 $state & P23147]
    (if (in-mem? $state M23145 P23147)
     ;; previously memoized result
     (fn []
       (C23144
         (get-mem $state M23145 P23147)
         $state))
     ;; new computation
     (clojure.core/apply
      (fn [C23150 $state person]
        (fn [] (C23150 ... $state)))
      ;; memoize result in state
      (fn [V23146 $state]
       (fn []
        (C23144
          V23146
          (set-mem $state M23145 P23147 V23146))))
      $state
      P23147))))
  $state))
\end{minted}
    \end{minipage}
\end{frame}

\begin{frame}[fragile]{Managing stack size}
    \begin{itemize}
        \item Clojure does not support tail-call optimization (TCO).
        \item Under CPS transformation the stack will explode.
            \pause
        \item Anglican uses \textit{trampolining}: every continuation
            call is wrapped into a \textit{thunk}.
    \end{itemize}

    The thunk is returned ...

    \begin{minipage}{0.48\textwidth}
\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
    (fn [x y] (+ x y))
\end{minted}
    \end{minipage}
    \begin{minipage}{0.48\textwidth}
        \vspace{0.5em}
\begin{minted}[linenos, fontsize=\scriptsize]{clojure}
    (fn [cont $state x y] 
      (fn []
        (cont (+ x y) $state)))
\end{minted}
      \vspace{0.5em}
    \end{minipage}

    ... and  called by \texttt{trampoline}:

\begin{minted}[linenos, fontsize=\small]{clojure}
        (defn exec
          [algorithm prog value state]
          (loop [step (trampoline prog value state)]
            (let [next (checkpoint algorithm step)]
              (if (fn? next)
                (recur (trampoline next))
                next))))
\end{minted}
